Copulas and their potential for ecology

### 2019-12-18 Paper.Rmd
working on final trim

<!--***DAN: remianing to do:
-->

@@

\noindent All branches of ecology study relationships among and between environmental and biological variables. However, 
standard approaches to studying such relationships, based on correlation and regression, 
provide only a small slice of the complex information contained in the relationships. Other 
statistical approaches exist that provide a complete description of relationships between variables, 
based on the concept of the *copula*; they are applied in finance, neuroscience and other fields, 
but rarely in ecology. We here explore the concepts that underpin copulas and examine the potential 
for those concepts to improve our understanding of ecology. We find that informative copula 
structure in dependencies between variables is common across all the environmental, 
species-trait, phenological, population, community, and ecosystem functioning datasets we 
considered. Many datasets exhibited asymmetric tail associations, whereby two variables were more 
strongly related in their left compared to right tails, or *vice versa*. We describe mechanisms by which 
observed copula structure and asymmetric tail associations can arise in ecological data, including a Moran-like 
effect whereby dependence structures between environmental variables are inherited by ecological 
variables; and asymmetric or nonlinear influences of environments on ecological variables, such as 
under Liebig's law of the minimum. We also describe consequences of copula structure for 
ecological phenomena, including impacts 
on extinction risk, Taylor's law, and the stability through time of ecosystem services. By documenting the 
importance of a complete description of dependence between variables, advancing conceptual frameworks, and 
demonstrating a powerful approach, we aim to encourage widespread use of copulas in ecology, which we
believe can benefit the discipline.

@@

All branches of ecology 
study relationships among biological variables and relationships between 
environmental and biological 
variables. However, commonly used correlation and regression 
approaches to studying 
such relationships are limited, and 
provide only a partial description of a relationship between variables. 
For instance, datasets showing markedly different 
relationships may have the same correlation coefficient
(Fig. \ref{fig_cop_pedag1}A-B). The variables of Fig. \ref{fig_cop_pedag1}A 
(respectively, Fig. \ref{fig_cop_pedag1}B) are principally 
related in the left (respectively, right) portions of their distributions, 
an asymmetric pattern of dependence that can
have ecological significance, as discussed below, but that is not captured by 
correlation. Although correlation is only one way to 
study relationships between variables, other common metrics also 
provide only partial information.

@@

Well-developed approaches do exist, however, and are 
applied widely in other fields, 
that provide a description of the dependence 
between two or more variables that can be mathematically proven 
to be complete [@nelsen2006_copula; @joe2014_dependence; @MaiScherer2017]. 
The approaches are based on 
the concept of the *copula*. Copula approaches separate the information of a bivariate 
random variable into two distinct parts: the information in the 
marginal distributions (which says nothing about the dependence between the 
variables), and the remaining information (which is solely about
dependence). We here introduce some simple concepts, based on taking ranks,
that relate to copulas and give a conceptual 
flavor of copula ideas to be introduced 
more formally in the "Background on copulas" section below (section \ref{Background}). Given 
a sample $(x_i,y_i)$, $i=1,2,\ldots,n$ (e.g., 
Fig. \ref{fig_cop_pedag1}A-B), information about the structure of the dependence
between $x$ and $y$ can be separated from information contained 
in the marginal distributions by 
considering the bivariate plot of $u_i$ versus $v_i$, where
$u_i$ is the rank of $x_i$ in the set $x_j$ ($j=1,\ldots,n$), divided by
$n+1$; and $v_i$ is defined in the same way but using the $y_i$. The rank
of the smallest element in a set is $1$. The $u_i$ and 
$v_i$ are called *normalized ranks* of the $x_i$ and $y_i$; they relate to 
the empirical cumulative distribution function of the $x_i$ and $y_i$,
respectively. We henceforth abbreviate "cumulative distribution function" as "cdf". 
See, for instance,
Fig. \ref{fig_cop_pedag1}C-D, which show the normalized ranks of 
A-B, respectively. Ranking makes the marginal distributions of 
the component datasets uniform, isolating the 
dependence structure. Dependence structure and marginals can then be studied 
separately. Normalized rank plots such as Fig. \ref{fig_cop_pedag1}C-D
relate to copulas in a way to be described in the "Background on copulas" section below
(section \ref{Background}), after a definition of copulas
is provided. Note that Pearson correlation, even though it is the most commonly
used measure of relationships between variables, is modified by monotonic
tranformation of the component variables (Appendix 
\ref{SM-pedagog1_details}, Fig. \ref{SM-fig_transf_pearson_spearman}),
and therefore reflects not only dependence information, but also
information on the marginals [@Genest2007]. 
Spearman and Kendall correlations, being rank-based, 
are not influenced by monotonic transformations of the component variables. 
They provide information solely on the dependence between those variables.

A main benefit of a copula approach is 
that it can detect associations in the tails of distributions, and asymmetries of such associations. 
Tail association (introduced formally in the "Background on copulas" section, below, section 
\ref{Background}) 
is association between extreme values of variables. 
If smaller values of two positively associated variables are more 
strongly associated than are larger values, the variables are said to
have stronger lower- or left-tail association than right- or upper-tail association
(Fig. \ref{fig_cop_pedag1}C);
and *vice versa* if larger values are more strongly associated than smaller
ones (Fig. \ref{fig_cop_pedag1}D). Datasets of the same 
Spearman or Kendall correlation can have a range of tail associations 
(Fig. \ref{SM-fig_cop_pedag}), which can be quantified (see sections on Background and
methods for Q1 below, sections \ref{Background} and \ref{Methods}). 
Although copulas can be used to model the entire
complex dependence structure of data, we have found tail associations to
usefully describe the full structure of dependence, so tail association 
is one of the main foci of this paper. We give four examples below which
illustrate some of the potential ecological meaning of tail associations.

The goal of this paper is to explore the potential for 
applications of copulas in ecology and to 
help determine to what extent ecological understanding may benefit generally from
using copulas. We provide a basic introduction
to copula concepts in the "Background on copulas" section below (section \ref{Background}). 
Readers can find additional abbreviated [@Anderson2018; @Genest2007] 
and comprehensive [@nelsen2006_copula; @joe2014_dependence; @MaiScherer2017]
introductions elsewhere. Copulas have already been used very effectively 
in a few studies to illuminate ecological phenomena 
[@Valpine2014; @Anderson2018; @Popovic2018], but such usage is
still rare. Whereas @Anderson2018 provide a specific methodological pathway, based partly 
on copulas, for the important problem of the 
analysis of multivariate ecological population count data;
we instead advocate more broadly for the application of copulas across ecology.
Our results suggest that environmental, 
ecological and evolutionary processes 
may commonly generate and transmit complex dependence structures, including asymmetric tail 
associations, and that greater use of copulas can illuminate the underlying processes.
We believe copula approaches are among the tools all ecologists, including tropical
ecologists, should be considering for analysis of their data in the 21st century.

@@

>>>>

@@

\noindent All branches of ecology study relationships among and between environmental and biological variables. However, 
standard approaches to studying such relationships, based on correlation and regression, 
provide only some of the complex information contained in the relationships. Other 
statistical approaches exist that provide a complete description of relationships between variables, 
based on the concept of the *copula*; they are applied in finance, neuroscience and elsewhere, 
but rarely in ecology. We explore the concepts that underpin copulas and the potential 
for those concepts to improve our understanding of ecology. We find that informative copula 
structure in dependencies between variables is common across all the environmental, 
species-trait, phenological, population, community, and ecosystem functioning datasets we 
considered. Many datasets exhibited asymmetric tail associations, whereby two variables were more 
strongly related in their left compared to right tails, or *vice versa*. We describe mechanisms by which 
observed copula structure and tail associations can arise in ecological data, including a Moran-like 
effect whereby dependence structures are inherited from environmental variables; and asymmetric or nonlinear influences of environments on ecological variables, such as 
under Liebig's law of the minimum. We also describe consequences of copula structure for 
ecological phenomena, including impacts 
on extinction risk, Taylor's law, and the temporal stability of ecosystem services. By documenting the 
importance of a complete description of dependence between variables, advancing conceptual frameworks, and 
demonstrating a powerful approach, we encourage widespread use of copulas in ecology, which we
believe can benefit the discipline.

@@

All branches of ecology 
study relationships among biological variables and relationships between 
environmental and biological 
variables. However, commonly used correlation and regression 
approaches to studying 
such relationships are limited, and 
provide only a partial description the relationship. 
For instance, datasets showing different 
relationships may have the same correlation coefficient
(Fig. \ref{fig_cop_pedag1}A-B). The variables of Fig. \ref{fig_cop_pedag1}A 
(respectively, Fig. \ref{fig_cop_pedag1}B) are principally 
related in the left (respectively, right) portions of their distributions, 
an asymmetric pattern of association that can
have ecological significance, as discussed below, but that is not captured by 
correlation. Although correlation is only one way to 
study relationships between variables, other common metrics also 
provide only partial information.

@@

Well-developed approaches do exist, however, and are 
applied widely in other fields, 
that provide a description of the association 
between two or more variables that is mathematically provably 
complete [@nelsen2006_copula; @joe2014_dependence; @MaiScherer2017]. 
The approaches are based on 
the concept of the *copula*. Copula approaches separate information in a bivariate 
random variable into two parts: the information in the 
marginal distributions (which says nothing about the association between the 
variables), and the remaining information (which is solely about
the association). We here introduce some simple concepts, based on ranks,
that relate to copulas and give a conceptual 
flavor of copula ideas to be introduced 
formally in the "Background" section below (section \ref{Background}). Given 
a sample $(x_i,y_i)$, $i=1,2,\ldots,n$ (e.g., 
Fig. \ref{fig_cop_pedag1}A-B), information about the structure of the association
between $x$ and $y$ can be separated from information contained 
in the marginal distributions by 
considering the plot of $u_i$ versus $v_i$, where
$u_i$ is the rank of $x_i$ in the set $x_j$ ($j=1,\ldots,n$), divided by
$n+1$; and $v_i$ is defined in the same way but using the $y_i$. The rank
of the smallest element in a set is $1$. The $u_i$ and 
$v_i$ are called *normalized ranks* of the $x_i$ and $y_i$; they relate to 
the empirical cumulative distribution function of the $x_i$ and $y_i$,
respectively. We henceforth abbreviate "cumulative distribution function" as "cdf". 
See, for instance,
Fig. \ref{fig_cop_pedag1}C-D, which show the normalized ranks of 
A-B. Ranking makes the marginal distributions of 
the component datasets uniform, isolating the 
dependence structure. Dependence structure and marginals can then be studied 
separately. Normalized rank plots such as Fig. \ref{fig_cop_pedag1}C-D
relate to copulas in a way to be described 
(section \ref{Background}). Note that Pearson correlation, even though it is the most commonly
used measure of association, is modified by monotonic
tranformation of the component variables (Appendix 
\ref{SM-pedagog1_details}, Fig. \ref{SM-fig_transf_pearson_spearman}),
and therefore reflects not only dependence information, but also
information on the marginals [@Genest2007]. 
Spearman and Kendall correlations, being rank-based, 
are not influenced by monotonic transformations of the variables. 
They provide information solely on the dependence between those variables.

A main benefit of a copula approach is 
that it can detect associations in the tails of distributions, and asymmetries of such associations. 
Tail association (introduced formally in section 
\ref{Background}) is association between extreme values of variables. 
If smaller values of two positively associated variables are more 
strongly associated than are larger values, the variables are said to
have stronger lower- or left-tail association than right- or upper-tail association
(Fig. \ref{fig_cop_pedag1}C);
and *vice versa* if larger values are more strongly associated 
(Fig. \ref{fig_cop_pedag1}D). Datasets of the same 
Spearman or Kendall correlation can have a range of tail associations 
(Fig. \ref{SM-fig_cop_pedag}), which can be quantified (see sections on Background and
methods for Q1 below, sections \ref{Background} and \ref{Methods}). 
Although copulas can be used to model the entire
complex dependence structure of data, we have found tail associations to
be a useful component of that structure, so tail association 
is a main focus of this paper. We give four examples below which
illustrate some of the ecological meaning of tail associations.

The goal of this paper is to explore the potential for 
applications of copulas in ecology and to 
estimate to what extent ecological understanding may benefit from
using copulas. We provide an introduction
to copula concepts in section \ref{Background}. 
Readers can find additional abbreviated [@Anderson2018; @Genest2007] 
and comprehensive [@nelsen2006_copula; @joe2014_dependence; @MaiScherer2017]
introductions elsewhere. Copulas have already been used effectively 
in a few studies to illuminate ecological phenomena 
[@Valpine2014; @Anderson2018; @Popovic2018], but such usage is
still rare. Whereas @Anderson2018 provide a specific method, based
on copulas, for important problems of the 
analysis of multivariate population count data,
we instead advocate more broadly for the application of copulas across ecology.
Our results suggest that environmental, 
ecological and evolutionary processes 
may commonly generate complex dependence structures, including asymmetric tail 
associations, and that greater use of copulas can illuminate underlying processes.
We believe copula approaches are among the tools all ecologists 
should be considering for analysis of their data in the 21st century.

@@

<!--***DAN: got to here with final trim-->
